---
title: "Final Exam  DS 520 Fall 2024"
author: "Charley Kirk"
date: "12/12/2024"
output:
  word_document: default
  html_document:
    df_print: paged
---


In general for this exam: 

All problems use the SleepStudy data set from the Lock5Data package. You should have installed this package at the beginning of the semester. If not, you'll need to install it now.

Graphs should be labeled well. 

Descriptions should be in context.

You may use your book (in print or on the Internet), your notes, the course Moodle page, and any R files you have on your machine. You may not use the Internet outside of the moodle page and your book. You may not ask for outside nor receive outside help other than asking your professor (or Dr. Roth if she is proctoring your session) if you have a question.

Library Load
```{r}
library(tidyverse)
library(Lock5Data) 
library(moderndive)
library(infer)
set.seed(1224)
```

The SleepStudy data set is from Lock5Data package that goes with Statistics Unlocking the Power of Data by Lock et. al. The data was collected as part of the following study:

Data source: Onyper, S., Thacher, P., Gilbert, J., and Gradess, S., “Class start times, sleep, and academic performance in college: A path analysis,” Chronobiology International, April 2012; 29(3):318–335.

Here are the variables we are considering from the data description

GPA-Grade point average (0-4 scale)

DASScore-Combined score for depression, anxiety and stress

Happiness-	Measure of degree of happiness (a score)

Stress-	Coded stress score: normal or high

AverageSleep-Average hours of sleep for all days

AllNighter-Had an all-nighter this semester? 1=yes, 0=no



You can use the code below to view a slice of the data set.
```{r}
slice(SleepStudy, 1:5)

```

1.  (2 points)From the package Lock5Data, read the details of the SleepStudy data file.The Details and Source section at the end. Based on these sections, what is the intended population? Note the details are a bit vague so there are multiple correct answers.

```{r}
?SleepStudy
```

The population is college students.

2. Consider the variable GPA:

a. (1 point) Find the mean GPA of the data. Name it point_est.


```{r}
point_est<-SleepStudy%>%
  summarise(mean(GPA))
point_est
```


b. (3 points) Make a bootstrap distribution for the mean GPA and visualize it.

```{r}
boot_dist<-SleepStudy%>%
  specify(response = GPA) %>%
  generate(reps=1000)%>%
  calculate(stat = "mean")
visualize(boot_dist)
```



c. (2 points) Could the bootstrap distribution be used to find a confidence interval for the mean GPA. Explain your answer.


Yes the bootstrap distribution could be used to find a confidence interval for the mean GPA. The bootstrap distribution acts as a simulation of the sampling distribution for the mean GPA so for example we could use the percentiles of the bootstrap distribution to generate a confidence interval since in theory they would approximate the percentiles of the sampling distribution.



d. (3 points) Regardless of your answr for part c create a 95% (normal) bootstrap confidence interval for the mean GPA and visualize it.

```{r}
boot_dist%>%
  get_confidence_interval(level = 0.95, type="percentile")
```


e. (4 points) Write a sentence interpreting your confidence interval in context.

We are 95% confident that the interval (3.191528, 3.291625) contains the true population mean for GPA.

f. (3 points) Make a 90% percentile bootstrap confidence interval for the mean GPA and visualize it.

```{r}
boot_dist%>%
  get_confidence_interval(level = 0.90, type="percentile")
```


g. (4 points) Which interval is wider the 90% or the 95% interval. Explain why your answer is true.

The 95% interval is wider. This is because intuitively we cannot be more confident without expanding the range of possible values.


3. (5 points) Compare and contrast the bootstrap distribution for mean GPA you made in 2 with the theoretical sampling distribution for the mean GPA for this population. What would be the same and what would be different?

The two distributions will have similar shapes, similar spreads and similar centers. Any of those three things could vary slightly between the two distributions.

4. Consider the variables Stress and AllNighter for this question.

a. (2 points) What is the code below doing? Why are we doing that? Make sure you run the code.



```{r}
SleepStudy=SleepStudy%>%mutate(AllNighter=as.factor(AllNighter))
View(SleepStudy)
```

This code is ensuring that AllNighter is treated as a categorical variable. We want it treated as categorical because the 1's and 0's in this case aren't meant to represent numerical quantities but rather they are meant to indicate whether or not a student had an all nighter.

b. (2 points) Find the difference in the proportion of students who had an all nighter for students with high levels of Stress - students with normal levels of Stress.

```{r}
SleepStudy%>%
  filter(AllNighter=="1")%>%
  count(Stress)
```


```{r}
prop_high=(8/34)
prop_normal=(26/34)
obs_diff_props<-prop_high-prop_normal
obs_diff_props
```


c. (2 points) Check the assumptions to use a theoretical interval and test for this parameter. Make sure to explain how you are a little worried here. However, you should proceed with testing in the subsequent questions. 

```{r}
SleepStudy%>%
  filter(AllNighter=="1")%>%
  count(Stress)
```

Although we don't know the full details of the dataset from the available R documentation, it seems reasonable to conclude each student in the study has exactly one row in the dataset based on the fact that the data was obtained from a formal study and the students were asked to complete survey questions, which probably wouldn't be given to the same student twice. So I would say the observations in this dataset are probably independent.

The "high" and "normal" categories have sizes of only 8 and 26 respectively which is kind of low. I suppose this is the thing to be worried about, other than the uncertainty of the independence assumption.

d. (2 points) Calculate the 90% theoretical confidence interval for the proportion of students who had an all nighter for students with high levels of Stress minus students with normal levels of Stress.

```{r}
#se=sqrt(((8/34)(1-(8/34))/34)+((26/34)(1-(26/34))/34))
#lower_ci<-obs_diff_props+1.64*se
#upper_ci<-obs_diff_props-1.64*se
#data.frame(lower_ci, upper_ci)
```

I don't think we learned the formula for computing the standard error for a difference in sample proportions. We discussed the formulas for computing the standard error for a mean and a proportion, and chapter 9 in the text talks about the standard error for a difference in sample means along with t tests, but we didn't cover the standard error for a difference in sample proportions. As for t testing, I don't know of any way to change the parameters for a t test in infer to proportion instead of mean, same with the t.test function. So the work above is an educated guess based on the standard error for a difference in sample means in chapter 9. The code isn't working though.

e. (4 points) Interpret your interval in context.



f. (5 points) The theoretical confidence interval and/or its code can be used to make a hypothesis test. Which hypothesis can you test and what significance level corresponds to the confidence level? State the null and alternative hypotheses. What conclusion can you make and why? Interpret your result of the hypothesis test in context. You may give a short interpretation in context rather than the longer scientific form interpretation and still receive full credit if it is complete and correct.




5. Consider the variables AverageSleep and AllNighter for this question.

a. (2 points)  Find the mean of average sleep for students who had an all nighter, the mean for students who did not have an all nighter, and the  difference between the mean average sleep hours for students who had an all nighter-students who did not.

```{r}
mean_allnight<-SleepStudy%>%
  filter(AllNighter=="1")%>%
  summarise(mean(AverageSleep))
mean_notallnight<-SleepStudy%>%
  filter(AllNighter=="0")%>%
  summarise(mean(AverageSleep))
obs_diff_means<-mean_allnight-mean_notallnight
mean_allnight
mean_notallnight
obs_diff_means
```


b. (3 points) Visualize AverageSleep and AllNighter together. What do you see? What do you wonder? Does it look like there is as significant difference in average hours of sleep for all nighter minus no all nighter?

```{r}
SleepStudy%>%
  ggplot(aes(AllNighter, AverageSleep))+
  geom_boxplot()
```
I see that the distribution of AverageSleep for students who have not had an all nighter looks more even than for students who have had an all nighter. While there may be a difference in the mean for AverageSleep for no all nighter and the mean for AverageSleep for all nighter I wouldn't call it statistically significant just from the graph. I wonder why the boxplot for students who did not have an all nighter has outliers while the boxplot for students who did have an all nighter does not.

c. (2 points) Write the null and alternative hypotheses for testing whether there is a difference in mean  average hours of sleep for all nighter minus no all nighter. Make sure to define your parameters.

$H_{0}:\mu_{A}=\mu_{N}$
$H_{A}:\mu_{A}\ne\mu_{N}$

Where $\mu_{A}$ is the mean of AverageSleep for students who did have an all nighter, $\mu_{N}$ is the mean of AverageSleep for students who did not have an all nighter, $H_{0}$ is the Null Hypothesis and $H_{A}$ is the Alternative Hypothesis.

d. (2 points) Generate the null distribution of the difference in average hours of sleep for all nighter minus no all nighter ($\delta$). 

```{r}
null_distribution_sleep <- SleepStudy %>% 
  specify(formula = AverageSleep ~ AllNighter) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("1", "0"))
visualize(null_distribution_sleep)
```


e. (2 points) Visualize how the observed difference compares to the null distribution of $\delta$.  Describe what this graph tells you.

```{r}
visualize(null_distribution_sleep)+
  shade_p_value(obs_stat = obs_diff_means, direction = "both")
```



f. (2 points) Calculate the p-value.

```{r}
null_distribution_sleep%>%
  get_p_value(obs_stat = obs_diff_means, direction = "both")
```

I think the reason a p-value of 0 is outputted here is because the exact p-value has only 0's up to a certain number of decimal places. So I suspect it's at least less than 0.00.


g. (4 points) State the conclusion of your hypothesis test in context. You may give a short interpretation in context rather than the longer scientific form interpretation and still receive full credit if it is complete and correct.

At the 5% significance level we reject the null hypothesis and conclude that there is a difference in the population mean of AverageSleep for students who have had an all nighter and the population mean of AverageSleep for students who have not had an all nighter

6. For the variables DASScore and Happiness.

a. (2 points) Graph DASScore and Happiness together. What type of relationship, if any, do you see between the two variables?

```{r}
SleepStudy%>%
  select(Happiness, DASScore)
```


```{r}
SleepStudy%>%
  ggplot(aes(DASScore, Happiness))+
  geom_point()
```


b. (2 points) Make a linear model for Happiness predicted by DASScore and get its regression table.


```{r}
happ_score_model<-lm(data=SleepStudy, Happiness~DASScore)
get_regression_table(happ_score_model)
```



c. (3 points) Looking at either the regression table or the summary for the model. Is the model valid? Explain your answer using a p-value and making a conclusion in context.

```{r}
summary(happ_score_model)
```

I would say that the model is valid. Since the p-value for the coefficient estimate of DASScore is below 0.05, at a 5% significance level I would conclude that there is a relationship between DASScore and Happiness.

d. (4 points) Looking at the regression table, give the 95% confidence interval for the slope and interpret it in context.

I am 95% confident that the interval (-0.196, -0.123) contains the true population slope relating DASScore and Happiness.


e. (2 points) Give an interpretation of the R-squared (r^2) value in the context of the model

For an R-squared value of 0.2314 we say that the model explains 23.14% of the variability in Happiness.


f. (4 points) Looking at context of the data and the graphs of the data and model, the residual plot, and the histogram of residuals. Does the model fit the LINE criteria? Explain.


```{r}
SleepStudy%>%
  ggplot(aes(DASScore, Happiness))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)
```

Since the relationship isn't obviously nonlinear I would say that this model meets the L condition of LINE.

Although we don't know the full details of the dataset from the available R documentation, it seems reasonable to conclude each student in the study has exactly one row in the dataset based on the fact that the data was obtained from a formal study and the students were asked to complete survey questions, which probably wouldn't be given to the same student twice. So I would say this dataset probably meets the I condition of LINE.

```{r}
happ_score_model%>%
  get_regression_points()%>%
  ggplot(aes(residual))+
  geom_histogram()
```
Although the histogram of the residuals contains some low outliers it appears as though it might otherwise be roughly normal shaped and centered on 0.

```{r}
happ_score_model%>%
  get_regression_points()%>%
  ggplot(aes(DASScore, residual))+
  geom_point()
```
Although the density of the points may appear to change a bit from left to right the actual distance from y=0 does not necessarily change dramatically. I would say the residuals do not obviously lack equal variance across the values of DASScore and so I would say the model meets the E condition of LINE.

g. (2 points) What is your assessment of the overall model? Would you use it? Why or why not?


While the model might not be invalid in that it could be set to meet LINE, that there is statistically significant evidence backing the existence of a relationship between DASScore and Happiness, and that it could be said to explain at least some of the variability in Happiness, I would probably not use the model. I would be more comfortable with an R-squared value higher than 0.2314.


