---
title: "InferenceProject"
author: "Charles Kirk"
date: "2024-11-21"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(infer)
library(tidyverse)
library(readr)
library(moderndive)
library(effsize)
```

```{r}
pres <- read_csv("~/Dropbox/Charley/Juniata/Data Science Masters/Statistics Fundamentals/CleanBloodpressuredata.csv")
pres<-pres%>%
  select(-REC, -Date)
```

Introduction

This dataset comes from an assignment in Data Science Fundamentals where we cleaned up a dataset of data on fair attendees. The data I used to create the presdat dataframe is a cleaned csv version of the Bloodpressuredata.xlsx document from that assignment. I believe the original source was given to be JC Blair Hospital, apparently renamed Penn Highlands Huntingdon. The hospital collected this data from people at the Huntingdon County Fair. The data may have been collected by survey but I don't think I have access to the original survey questions. There may be a lot of NAs in the data that get filtered out as I perform hypothesis tests and construct confidence intervals, so I might not be working with subsets that are super representative of the population. The population of interest is the attendees of the Huntingdon County Fair.

My CleanBloodpressuredata csv contained 19 columns but I selected just 17 for this project, omitting REC and Date.

Variables: OVERALL: Case ID variable. YEAR: The year the individual's data was recorded. BP: The blood pressure of the individual; I think it's computed as Systolic/Diastolic though it may not be directly derived from the Systolic and Diastolic in this dataset. BPRatio:The individual's BP given as a decimal. Systolic: In the HospitalDataCleaning.Rmd document it was described as "the top number in blood pressure." Diastolic: In the HospitalDataCleaning.Rmd document it was described as "the bottom number in blood pressure." BPStatus: Low, Normal, Pre-Hypertension or Hypertension. Age: The age of the individual in years. Weight: The weight of the individual in pounds. HeightInches: The height of the individual in inches. BMI: Computed based on the height and weight of the individual. BMICategory: Normal, Overweight or Obese. Gender: The gender of the individual, with values of M and F. DiagnosedDiabetic: Y if the individual was ever diagnosed with diabetes, N if they were not. TobaccoUser: Y if the individual uses tobacco, N if they do not. OnmedicationforBP: Y if the individual is on medication for high blood pressure, N if they are not. HuntingdonCountyResident: Y if the individual is a resident of Huntingdon County, N if they are not.

Here I perform some additional cleaning of my data.
```{r}
pres<-pres%>%
  mutate(DiagnosedDiabetic=fct_recode(DiagnosedDiabetic, Y="Gestational Diabetes", Y="During Pregnancy", N="Borderline", N="Pre-Diabetic"), BPStatus=fct_recode(BPStatus, NULL="N/A", NULL="NA", Other="Optimal"), OnmedicationforBP=fct_recode(OnmedicationforBP, NULL="N/A"), TobaccoUser=fct_recode(TobaccoUser, NULL="N/A", N='M'))%>%
  filter(is.na(OnmedicationforBP)==TRUE | OnmedicationforBP!='128')
pres
```


Confidence Intervals

Question 1: Is there a significant difference in the average age of attendees who are diagnosed with diabetes and attendees who are not diagnosed with diabetes? To answer this question I will construct a 90% confidence interval for the difference in population means average age of attendees diagnosed with diabetes and average age of attendees not diagnosed with diabetes using the bootstrap percentile method.

```{r}
pres%>%
  filter(is.na(DiagnosedDiabetic)==FALSE)%>%
  count(DiagnosedDiabetic)
pres%>%
  summarise(mean(Age, na.rm = TRUE))
fivenum(pres$Age)
```

```{r}
pres%>%
  ggplot(aes(DiagnosedDiabetic, Age))+
  geom_boxplot()
```


```{r}
set.seed(35)
bootpres<-pres%>%
  filter(is.na(DiagnosedDiabetic)==FALSE)%>%
  specify(formula=Age~DiagnosedDiabetic)%>%
  generate(reps=1000, type="bootstrap")%>%
  calculate(stat = "diff in means", order = c("Y", "N"))
bootpres
```

```{r}
bootpres%>%
  get_confidence_interval(level=0.9, type="percentile")
```
We used the 5th and 95th percentiles of the bootstrap distribution to construct a confidence interval. This method is appropriate because the bootstrap distribution was created with bootstrap resampling with replacement to be a simulation of the sampling distribution for the difference in means. We have 90% confidence that the interval (5.879943, 8.058591) contains the true difference in population means $\mu_{Y} - \mu_{N}$.


Question 2: Roughly what proportion of attendees have a “normal” blood pressure status? To answer this question I will construct a 95% confidence interval for the population proportion of attendees with a “normal” blood pressure status using the theoretical standard error method.

```{r}
pres%>%
  count(BPStatus)
```

```{r}
pres%>%
  mutate(BPStatus=fct_recode(BPStatus, Other='Hypertension', Other='Pre-Hypertension'))%>%
  ggplot(aes(BPStatus))+
  geom_bar()
```


```{r}
663/(900+1023+637+663)
```


```{r}
proportion<-pres%>%
  mutate(BPStatus=fct_recode(BPStatus, Other='Hypertension', Other='Pre-Hypertension'))%>%
  filter(is.na(BPStatus)==FALSE)%>%
  specify(response = BPStatus, success = "Normal")%>%
  calculate(stat = "prop")
proportion
```

```{r}
se=sqrt((proportion*(1-proportion))/(900+1023+637+663))
lower_ci<-proportion-1.96*se
upper_ci<-proportion+1.96*se
data.frame(lower_ci, upper_ci)%>%
  rename(lower_ci=stat, upper_ci=stat.1)
```
This method is appropriate because the expression sqrt((proportion*(1-proportion))/(900+1023+637+663)) gives an estimate of the standard error of the sampling distribution for the sample proportion, and it can be shown that 95% of values are within +-1.96 standard errors of the sample proportion. We have 95% confidence that the interval (0.1917536, 0.2196644) contains the true population proportion $p_{Normal}$.

Question 3: Is there a significant difference in the average BMI of attendees who are male and attendees who are female? To answer this question I will construct a 95% confidence interval for the difference in population means average BMI of attendees who are male and average BMI of attendees who are female using the bootstrap standard error method.

```{r}
pres%>%
  count(Gender)
pres%>%
  filter(is.na(BMI))
pres%>%
  summarise(mean(BMI, na.rm = TRUE))
fivenum(pres$BMI)
```


```{r}
pres%>%
  ggplot(aes(Gender, BMI))+
  geom_boxplot()
```


```{r}
set.seed(84)
bootpres2<-pres%>%
  filter(is.na(Gender)==FALSE)%>%
  specify(formula=BMI~Gender)%>%
  generate(reps=1000, type="bootstrap")%>%
  calculate(stat = "diff in means", order = c("M", "F"))
```

```{r}
bootpres2%>%
  ggplot(aes(stat))+
  geom_histogram()
```
The distribution looks relatively bell-shaped.


```{r}
point_estimate<-pres%>%
  filter(is.na(Gender)==FALSE)%>%
  specify(formula=BMI~Gender)%>%
  calculate(stat = "diff in means", order = c("M", "F"))
point_estimate
```


```{r}
bootpres2%>%
  get_confidence_interval(level=0.95, type="se", point_estimate = point_estimate)
```
We used the standard error of the bootstrap distribution to construct a confidence interval. This method is appropriate because the bootstrap distribution was created with bootstrap resampling with replacement to be a simulation of the sampling distribution for the difference in means, and because the distribution is relatively bell-shaped we can typically expect around 95% of values to be within +-1.96 standard errors of the point estimate. We have 95% confidence that the interval (-0.04248819, 1.06699) contains the true difference in population means $\mu_{M} - \mu_{F}$.


Hypothesis Tests

Question 1: Is there a significant difference in the average BP Ratio of attendees who are on medication for BP and attendees who are not on medication for BP? I will answer this question by conducting a hypothesis test using a 5% significance level and a null distribution constructed through randomization of my original sample.

```{r}
pres%>%
  count(OnmedicationforBP)
pres%>%
  summarise(mean(BPRatio, na.rm = TRUE))
fivenum(pres$BPRatio)
```


```{r}
pres%>%
  ggplot(aes(OnmedicationforBP, BPRatio))+
  geom_boxplot()
```



```{r}
set.seed(45)
obs_diff_means<-pres %>%
  filter(is.na(OnmedicationforBP)==FALSE)%>%
  specify(BPRatio ~ OnmedicationforBP) %>%
  calculate(stat = "diff in means", order = c("Y", "N"))
null_distribution_BPRatio <- pres %>%
  filter(is.na(OnmedicationforBP)==FALSE)%>%
  specify(BPRatio ~ OnmedicationforBP)%>%
  hypothesize(null = "independence") %>% 
  generate(reps = 5000)%>%
  calculate(stat = "diff in means", order = c("Y", "N"))
null_distribution_BPRatio%>%
  get_p_value(obs_stat = obs_diff_means, direction = "both")
obs_diff_means
```

The mean BP Ratio of attendees who are on medication for BP in my sample was greater than the mean BP Ratio of attendees who are not on medication for BP in my sample. The p-value for this test was obtained by checking the observed difference in means for the two medication groups against a distribution that was obtained by generating samples of the original dataset with the values of OnmedicationforBP shuffled. This is appropriate because said distribution acts as a null distribution where the difference in means is 0. The p-value was approximated at 0 which is less than $\alpha = 0.05$. Given this I would reject the Null hypothesis and conclude that attendees who are on medication for BP have a higher BP Ratio than attendees who are not on medication for BP.

Question 2: Is there a significant difference in the average weight of attendees who are on medication for BP and attendees who are not on medication for BP? I will answer this question by conducting a hypothesis test using a 5% significance level and a null distribution constructed through randomization of my original sample.

```{r}
pres%>%
  summarise(mean(Weight, na.rm = TRUE))
fivenum(pres$Weight)
```


```{r}
pres%>%
  ggplot(aes(OnmedicationforBP, Weight))+
  geom_boxplot()
```


```{r}
set.seed(83)
obs_diff_means<-pres %>%
  filter(is.na(OnmedicationforBP)==FALSE)%>%
  specify(Weight ~ OnmedicationforBP) %>%
  calculate(stat = "diff in means", order = c("Y", "N"))
null_distribution_Weight <- pres %>%
  filter(is.na(OnmedicationforBP)==FALSE)%>%
  specify(Weight ~ OnmedicationforBP)%>%
  hypothesize(null = "independence") %>% 
  generate(reps = 5000)%>%
  calculate(stat = "diff in means", order = c("Y", "N"))
null_distribution_Weight%>%
  get_p_value(obs_stat = obs_diff_means, direction = "both")
obs_diff_means
```

The mean weight of attendees who are on medication for BP in my sample was greater than the mean weight of attendees who are not on medication for BP in my sample. The p-value for this test was obtained by checking the observed difference in means for the two medication groups against a distribution that was obtained by generating samples of the original dataset with the values of OnmedicationforBP shuffled. This is appropriate for the same reason as in the last hypothesis test, because said distribution acts as a null distribution where the difference in means is 0. The p-value was approximated at 0 which is less than $\alpha = 0.05$. Given this I would reject the Null hypothesis and conclude that attendees who are on medication for BP have a higher weight than attendees who are not on medication for BP.


Question 3: Is there a significant difference in the average age of attendees who are tobacco users and attendees who are not tobacco users? I will use a two-sample t-test and a 5% significance level to answer this question.

```{r}
pres%>%
  count(TobaccoUser)
pres%>%
  summarise(mean(Age, na.rm = TRUE))
fivenum(pres$Age)
```

```{r}
pres%>%
  ggplot(aes(TobaccoUser, Age))+
  geom_boxplot()
```

We will check for normality and equal variance.

```{r}
pres%>%
  filter(TobaccoUser=="Y")%>%
  ggplot(aes(Age))+
  geom_histogram()
```


```{r}
pres%>%
  filter(TobaccoUser=="N")%>%
  ggplot(aes(Age))+
  geom_histogram()
```

The histograms seem relatively bell-shaped.

```{r}
pres %>% 
  filter(is.na(TobaccoUser)==FALSE)%>%
  group_by(TobaccoUser) %>% summarise("standard deviation"=sd(Age, na.rm=TRUE))
```
The larger standard deviation is not more than twice the smaller standard deviation.

```{r}
cohen.d(Age~ TobaccoUser, data = pres,  paired = FALSE,pooled=TRUE,conf.level=0.95)
```


```{r}
t.test(Age ~ TobaccoUser, data = pres, 
       alternative = "two.sided", 
       mu = 0, 
       var.equal = TRUE,  
       conf.level = 0.95)
```

I used a two-tailed independent (unpaired) t test to test the Null Hypothesis that there was no relationship between tobacco use of an attendee and the age of an attendee **at the population level**. The age was greater for attendees who did not use tobacco  ( x-bar = 57.45070, *SD* = 17.32171) than for attendees who did use tobacco ( x-bar = 51.23226, *SD* = 14.84427). The difference between means was small (Cohen's *d* = 0.3671545). The p-value is 4.597e-16 and is less than $\alpha$ = 0.05 (*t* = 8.1599, *df* = 3572). There was enough evidence to reject the Null Hypothesis. I conclude from this study that there is a relationship between tobacco use of an attendee and the age of an attendee at the population level.

Inference for Regression

```{r}
HeightInches_Age<-lm(HeightInches ~ Age, data = pres)
```
I will graph this model and check whether it meets the conditions of LINE.

```{r}
pres%>%
  ggplot(aes(Age, HeightInches))+
  geom_point()+ 
  geom_smooth(method = "lm", se = FALSE)
```
The relationship between Age and HeightInches does not appear to be obviously nonlinear.

Regarding the independence condition of LINE, each row in the pres dataframe should in theory represent a unique attendee of the Huntingdon County Fair.

```{r}
get_regression_points(HeightInches_Age)%>%
  ggplot(aes(residual))+
  geom_histogram()
```
This histogram seems to be relatively bell-shaped and centered on 0.

```{r}
get_regression_points(HeightInches_Age)%>%
  ggplot(aes(Age, residual))+
  geom_point()
```
This first residual plot appears to have a relatively constant spread of residuals around the line y=0.

```{r}
get_regression_points(HeightInches_Age)%>%
  ggplot(aes(HeightInches, residual))+
  geom_point()
```
This second residual plot appears to show a strong linear trend. This could suggest that the Age variable doesn't fully capture the variation in HeightInches.

```{r}
get_regression_table(HeightInches_Age)
summary(HeightInches_Age)
```
The intercept doesn't really have a practical interpretation in this context so we can ignore it's p-value. The p-value for the coefficient of Age is calculated as 0.000481, so at a 5% significance level we reject the null hypothesis that there is no relationship between Age and HeightInches and conclude that there is a relationship between Age and HeightInches. Also based on the values given in the lower_ci and upper_ci columns in the regression table output we can say that we are 95% confident that the interval (-0.025, -0.007) contains the true population slope $\beta_{1}$. This interval notably doesn't include 0. But while I wouldn't necessarily say that this model is invalid, I wouldn't consider it the best model to use considering the fact that there is still a trend in the residuals that is not explained by Age and the multiple R-squared value is low at 0.004272.

Conclusion

I think a lot of my results turned out to be statistically significant, which is maybe a little surprising. It was interesting to observe that there was a statistically significant difference in the age of tobacco users and the age of tobacco users for example. That said I did unfortunately have a lot of NAs in my data so it wasn't a perfect sample. Regarding the linear model I created, I wasn't quite sure what to make of the second residual plot. Even if the model was a bad fit, I'm not sure why I would observe such a strong positive linear trend in the residuals like that. Maybe there's another variable that explains a lot of the variation in HeightInches, one that might not even be present in the dataset?